An Open-Source Implementation of the CMPS Algorithm for Assessing  Similarity of Bullet Signatures

### 2021-04-01 CMPSpaper_writing/CMPSpaper_writing.Rmd
steps 1-3

---
title: Capitalized Title Here
header-includes:
  - \usepackage{xcolor}
  - \usepackage{longtable}
author:
  # see ?rjournal_article for more information
  - name: Author One
    affiliation: ISU, CSAFE
    address:
    - line 1
    - line 2
    url: https://journal.r-project.org
    orcid: 0000-0002-9079-593X
    email:  author1@work
  - name: Heike Hofmann
    url: https://journal.r-project.org
    email: hofmann@iastate.edu
    orcid: 0000-0002-9079-593X
    affiliation: ISU, CSAFE
    address:
    - line 1 affiliation 1
    - line 2 affiliation 1
    affiliation2: Affiliation 2
    address2:
    - line 1 affiliation 2
    - line 2 affiliation 2
abstract: >
  An abstract of less than 150 words.
preamble: |
  % Any extra LaTeX you need in the preamble

@@

The automated framework  used in this paper was proposed by @aoas. In order to obtain bullet signatures from the x3p files of bullet land engravings, we use the `R` packages `x3ptools` [@x3ptools] and `bulletxtrctr` [@bulletxtrctr]. `x3ptools` \hh{is a package to read, write, and generally, process } x3p files. 
<!--In the case of Hamby set 252, we need to convert the measurements of x3p files from meters to microns (`x3ptools::x3p_m_to_mum`) and rotate x3p files to make the long axis horizontal (`x3ptools::x3p_rotate`).--> 
<!--\hh{it's land **engraved** areas, not impressions (that was wrong in Eric's first paper). Impressions are marks from impacts, such as the hit of a hammer or the hit from the breech face. Engravings are scratch marks.}-->
\hh{\autoref{fig:process} gives an overview of all of the steps in the process from scan to signatures.}
\autoref{fig:process}(a) shows a \hh{rendering of a 3d scan of a } bullet land  engraved area. The raised portion of the surface \hh{on the left and right of the scan are parts of the adjacent groove engraved areas (GEAs)}, \hh{the middle area shows a land-engraved area (LEA) with well expressed} striation marks. The first step of obtaining the bullet signature is to extract a cross-section at a fixed height along the land engraving.
<!--\hh{The next section is at an awkward level of detail - there's lots of details, but without the parameter settings it is still not a reproducible description. Could you instead describe things conceptually, and then link to the code that does this particular image?} -->
The white horizontal line in \autoref{fig:process}(a) indicates which cross-section was identified by the algorithm to represent the LEA; \autoref{fig:process}(b) shows the corresponding cross-sectional view. <!--Then we can acquire the crosscut data (`ccdata`) of the scan at this height. --> \hh{XXX passive voice for things that are not part of this paper :)}
Groove engraved areas are removed from the analysis as indicated by the vertical blue lines in \autoref{fig:process}(c) and \autoref{fig:process}(d)). A non-parametric LOESS smooth  [@loess] is fit to capture the bullet curvature (\autoref{fig:process}(e)) and, finally, the **bullet signature** (\autoref{fig:process}(f)) is obtained as residuals of the cross-section and the fitted smooth. 
\hh{Identifiying the groove engraved areas correctly is fundamental for a correct down-stream analysis of the signatures. To allow for a human-in-the middle inspection and intervention we have provided an interactive web-application, implemented as `R` Shiny App [@shiny],}   `bulletinspectR` to identify and correct those errors. 
\hh{An example of the extraction process with corresponding code and parameter settings can be found at XXX.}
Note that \hh{the process of extracting signatures might be different from the one used in @cmps because no code or parameter settings are made available publicly.}

@@

```{r bullet, echo=FALSE, out.width=".8\\textwidth", fig.cap="image of a XXX gun barrel and a fired bullet - \\hh{xxx my version of the image still has a groove engraved area marked as a land}"}
knitr::include_graphics("img/barrel_bullet_ps.png", dpi = 100)
```

@@

### Conceptual idea of CMPS
\hh{Most algorithms for comparing striation marks are based on the digitized signatures XXX and produce a similarity score (some citations: chumbley score, Ganesh's fix, vorburger's ccf paper, hare).} 
 The congruent matching profile segments (CMPS) algorithm, developed by @cmps for "objective comparison of striated tool marks", is one such algorithm. The algorithm's main idea  is to take a set of consecutive and non-overlapping basis segments from  the reference profile  and  for each segment find the "best" registration position on the comparison profile (the other bullet signature) with respect to their cross-correlation values. \hh{From a comparison of these registration positions, a } **congruent registration position** is identified, and the number of basis segments taking the congruent registration position is the CMPS score. 
\hh{High CMPS scores are achieved between more similar signatures and are therefore indicative of a same-source pair. Low scores between pairs of signatures are attributed to different source pairs.} 
\hh{XXX What threshold did @cmps identify as the boundary between different source and same source? And what was the error rate in their example?}
 The CMPS algorithm can assist firearm examiners with drawing a conclusion \hh{about the source of a comparison pair}. \hh{Unfortunately, @cmps did not release any code or specific parameter settings for their implementation of the } CMPS algorithm. 
Thus, in this paper we present an open-source implementation of the CMPS algorithm \hh{in the R package CMPS available from XXX github / CRAN}. 
With our implementation, one can compute the CMPS score of a comparison using the following code:

```{r, eval=FALSE}
cmps.result <- extract_feature_cmps(sig1, sig2)
```

\hh{XXX How about a barchart with results - or a teaser for an image in the back, that talks about the concept that a CMPS score should be high for two same-source signatures and low for different-source signatures}

@@

1. **cut the reference profile into consecutive, non-overlapping and equal-length basis segments:**

2. **register each basis segment on the comparison profile:** 
Calculate the cross-correlation curve between a basis segment and the comparison profile.A registration position is considered as "good" if the basis segment achieves a high cross-correlation value at that position. 
The best registration position is decided upon using  a multi-peak inspection with or without different segment lengths:

    2.1. **multi-peak inspection (without different segment lengths):** each basis segment identifies positions of 5 (or multiple) highest peaks in the cross-correlation curve. Since false positive peaks would stand out when the segment length is short, positions of all 5 peak will be considered as the potential "best" registration position.

    2.2. **multi-peak inspection with different segment lengths:** for this version of the CMPS algorithm, each basis segment finds only one "best" registration position. This "best" registration position can be found by conducting multi-peak inspection multiple rounds. The first round is the standard way as we have stated in step 2.1. But after the first round, we will increase the segment length, compute a new cross-correlation curve, and again identify multiple peaks in the cross-correlation curve. The number of peaks identified in this round is typically less than that in the previous round (say 3 highest peaks) since increasing the segment length can reduce the number of false positive peaks. We can repeat this procedure multiple times; however in the last round, only one highest peak should be identified. If a position can be identified for all different segment length, this position will be selected as the "best" registration position and is called the "consistent correlation peak" (ccp).

3. **determine the congruent registration position** with a small tolerance zone and the number of basis segments that take the congruent registration position as their "best" registration position. These segments are called the "congruent matching profile segments" (CMPS), and the number of CMPS is the CMPS score of this profile comparison.

@@

* The command `get_segs(x, len=50)` implements step 1:  It takes a profile {XXX profile or signature?} in the format of a numeric vector and cuts it into consecutive, non-overlapping and equal-length segments of length `len`.

* `get_ccf4` computes the cross-correlation function / correlation coefficient curve between two profiles. This is the function called whenever we need to compute the ccf.

@@

* `get_CMPS` considers all basis segments and their identified peak positions to determine the congruent registration position. The goal is to find a peak position that is agreed by the most basis segments. Since small errors are allowed, a tolerance zone would serve as a selection window. `get_CMPS` will move this selection window from left to right and go through all possible positions and check the number of basis segment that identifies a peak in the selection window for each position. Since we are moving this tolerance zone from left to right one unit a time, some consecutive positions are expected to have the same number of basis segments. [?.?]In that case the median of all tied and consecutive positions will be chosen as the congruent registration position. As long as the congruent registration position is determined, the number of basis segments that identify a peak in this position is the CMPS score.

>>>>

---
title: An Open-Source Implementation of the CMPS Algorithm for Assessing  Similarity of Bullet Signatures
header-includes:
  - \usepackage{xcolor}
  - \usepackage{longtable}
author:
  # see ?rjournal_article for more information
  - name: Will W. Ju
    affiliation: Department of Statistics
    address:
    - Center for Statistics and Applications in Forensic Evidence
    - Iowa State University
    - 2438 Osborn Dr
    - Ames, IA 50011
    url: https://github.com/willju-wangqian
    orcid: 0000-0002-9079-593X
    email:  wju@iastate.edu
  - name: Heike Hofmann
    url: https://github.com/heike
    email: hofmann@iastate.edu
    orcid: 0000-0002-9079-593X
    affiliation: Department of Statistics
    address:
    - Center for Statistics and Applications in Forensic Evidence
    - Iowa State University
    - 2438 Osborn Dr
    - Ames, IA 50011
abstract: >
  An abstract of less than 150 words.
preamble: |
  % Any extra LaTeX you need in the preamble

@@

The automated framework  used in this paper was proposed by @aoas. In order to obtain bullet signatures from the x3p files of bullet land engravings, we use the `R` packages `x3ptools` [@x3ptools] and `bulletxtrctr` [@bulletxtrctr]. `x3ptools` \hh{is a package to read, write, and generally, process } x3p files. \hh{The `bulletxtrctr` package implements a pipeline for extracting and comparing signatures from scans of land-engraved areas.}
<!--In the case of Hamby set 252, we need to convert the measurements of x3p files from meters to microns (`x3ptools::x3p_m_to_mum`) and rotate x3p files to make the long axis horizontal (`x3ptools::x3p_rotate`).--> 
<!--\hh{it's land **engraved** areas, not impressions (that was wrong in Eric's first paper). Impressions are marks from impacts, such as the hit of a hammer or the hit from the breech face. Engravings are scratch marks.}-->
\hh{\autoref{fig:process} gives an overview of all of the steps in the process from scan to signatures.}
\autoref{fig:process}(a) shows a \hh{rendering of a 3d scan of a } bullet land  engraved area. The raised portion of the surface \hh{on the left and right of the scan are parts of the adjacent groove engraved areas (GEAs)}, \hh{the middle area shows a land-engraved area (LEA) with well expressed} striation marks. The first step of obtaining the bullet signature is to extract a cross-section at a fixed height along the land engraving.
<!--\hh{The next section is at an awkward level of detail - there's lots of details, but without the parameter settings it is still not a reproducible description. Could you instead describe things conceptually, and then link to the code that does this particular image?} -->
The white horizontal line in \autoref{fig:process}(a) indicates which cross-section was identified by the algorithm to represent the LEA; \autoref{fig:process}(b) shows the corresponding cross-sectional view. <!--Then we can acquire the crosscut data (`ccdata`) of the scan at this height. --> \hh{XXX passive voice for things that are not part of this paper :)}
Groove engraved areas are removed from the analysis as indicated by the vertical blue lines in \autoref{fig:process}(c) and \autoref{fig:process}(d)). A non-parametric LOESS smooth  [@loess] is fitted to capture the bullet curvature (\autoref{fig:process}(e)) and, finally, the **bullet signature** (\autoref{fig:process}(f)) is obtained as residuals of the cross-section and the fitted smooth. 
\wj{Note that in @cmps bullet signatures are sometimes referred as bullet profiles. However, to avoid confusion, we distinguish the notion of bullet signatures from bullet profiles. Bullet profiles are XXX, while bullet signatures are XXX.}
\hh{Identifiying the groove engraved areas correctly is fundamental for a correct down-stream analysis of the signatures. To allow for a human-in-the middle inspection and intervention we have provided an interactive web-application, implemented as `R` Shiny App [@shiny],}   `bulletinspectR` to identify and correct those errors. 
\hh{An example of the extraction process with corresponding code and parameter settings can be found at XXX.}
Note that \hh{the process of extracting signatures might be different from the one used in @cmps because no code or parameter settings are made available publicly.}

@@

```{r bullet, echo=FALSE, out.width=".8\\textwidth", fig.cap="Photo of a traditionally rifled gun barrel (left) and a fired bullet (right)."}
knitr::include_graphics("img/barrel_bullet_ps.png", dpi = 100)
```

@@

### Conceptual idea of CMPS
\hh{Most algorithms for comparing striation marks are based on the digitized signatures XXX and produce a similarity score (some citations: chumbley score, Ganesh's fix, vorburger's ccf paper, hare).} 
 The congruent matching profile segments (CMPS) algorithm, developed by @cmps for "objective comparison of striated tool marks", is one such algorithm. The algorithm's main idea  is to take a set of consecutive and non-overlapping basis segments from  the reference profile \wj{might reference profile refer to something more general?}  and  for each segment find the "best" registration position on the comparison profile (the other bullet signature) with respect to their cross-correlation values. \hh{From a comparison of these registration positions, a } **congruent registration position** is identified, and the number of basis segments taking the congruent registration position is the CMPS score. 
\hh{High CMPS scores are achieved between more similar signatures and are therefore indicative of a same-source pair. Low scores between pairs of signatures are attributed to different source pairs.} 
\hh{XXX What threshold did @cmps identify as the boundary between different source and same source? And what was the error rate in their example?}
\wj{they didn't talk about a specific threshold or error rates}
 The CMPS algorithm can assist firearm examiners with drawing a conclusion \hh{about the source of a comparison pair}. \hh{Unfortunately, @cmps did not release any code or specific parameter settings for their implementation of the } CMPS algorithm. 
Thus, in this paper we present an open-source implementation of the CMPS algorithm \hh{in the R package CMPS available from XXX github / CRAN}. 
With our implementation, one can compute the CMPS score of a comparison using the following code:

\hh{XXX could you make this example functional?}
```{r, eval=FALSE}
cmps.result <- extract_feature_cmps(sig1, sig2)
```

\hh{XXX How about a barchart with results - or a teaser for an image in the back, that talks about the concept that a CMPS score should be high for two same-source signatures and low for different-source signatures}
\wj{further question?}

@@

1. **cut the reference profile into consecutive, non-overlapping and equal-length basis segments:**
The command `get_segs(x, len=50)` implements this step:  it takes a bullet signature  in the format of a numeric vector and cuts it into consecutive, non-overlapping and equal-length segments of length `len`.

2. **register each basis segment on the comparison profile:** 
Calculate the cross-correlation curve between a basis segment and the comparison profile (using function `get_ccf4`).
A position is considered  "good" if the cross-correlation between the basis segment and the comparison is above a pre-defined threshold. \hh{XXX what is the name of the parameter?} 
The best registration position is decided upon using  a multi-peak inspection with or without different segment lengths:

    2.1. **multi-peak inspection (without different segment lengths):** each basis segment identifies positions of the top X \hh{parameter name?} highest peaks in the cross-correlation curve. Small segment lengths might indicate false positive peaks. Therefore all X peaks are considered as candidates for the "best" registration position.

    2.2. **multi-peak inspection with different segment lengths:** for this version of the CMPS algorithm, each basis segment finds only one "best" registration position. This "best" registration position can be found by conducting multi-peak inspection multiple rounds. The first round is executed as stated in step 2.1. After that first round, the segment length is increased, a new cross-correlation curve is computed, and another set of  peaks in the cross-correlation curve is identified together with the candidate positions. Typically, fewer number of peaks are identified in this round (because the increased segment length reduces the chance of false positive peaks). This process can be repeated multiple times. In the last round, only one highest peak should be identified. If the same or 'similar enough' (see step 3) position can be identified for all different segment length, this position will be selected as the "best" registration position and is called the "consistent correlation peak" (ccp).

3. **determine the congruent registration position**:
candidate positions across different segements are compared using a small tolerance \hh{XXX parameter for this} and basis segments with a congruent registration position are determined. These segments are called the "congruent matching profile segments" (CMPS), and the number of CMPS is the CMPS score of the comparison.

@@

@@

* `get_CMPS` considers all basis segments and their identified peak positions to determine the congruent registration position. The goal is to find a peak position that is agreed by the most basis segments. Since small errors are allowed, a tolerance zone would serve as a selection window. `get_CMPS` will move this selection window from left to right and go through all possible positions and check the number of basis segment that identifies a peak in the selection window for each position. Since we are moving this tolerance zone from left to right one unit a time, some consecutive positions are expected to have the same number of basis segments. [?.?] In that case the median of all tied and consecutive positions will be chosen as the congruent registration position. As long as the congruent registration position is determined, the number of basis segments that identify a peak in this position is the CMPS score.
